```{r page_setup_excise, include=FALSE}
options(encoding = "UTF-8")
try({
  if (.Platform$OS.type == "windows") {
    Sys.setlocale("LC_CTYPE", "English_United States.utf8")
  } else {
    Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
  }
}, silent = TRUE)
```

# Excise Module

The Excise Module estimates baseline excise duty liabilities and simulates the fiscal, price, and volume effects of changes in excise policy parameters. The current implementation focuses on alcohol products: Beer, Spirits, and Wine. It is designed to show how an alcohol excise model can combine product volumes, retail prices, excise rates, VAT rates, tariffs, alcohol-by-volume information, macroeconomic assumptions, elasticities, and retail margins to estimate baseline and reform-scenario outcomes.

The module uses artificial synthetic input data generated directly in R. The Shiny application imports the generated model input file at startup and then runs the excise calculation scripts inside the same R session. The synthetic data are designed for demonstration and training purposes. They reproduce the structure required by the model, but they do not represent official excise returns, taxpayer declarations, customs data, market data, or a specific country.

The Excise Module estimates business-as-usual outcomes and compares them with a simulation scenario. Users can choose the simulation year, select an excise product, choose the type of excise, enter a simulated excise rate, and optionally apply the change to following years. Users can also adjust price elasticity assumptions, income elasticity, and the retail margin. After the simulation is run, the module produces volume tables, nominal excise revenue tables, real excise revenue tables, and dashboard visualizations.

## Required excise data

The Excise Module requires input objects that describe the alcohol excise base, the applicable tax rates, the policy structure, behavioural assumptions, and macroeconomic assumptions. In the current Shiny implementation, these objects are loaded from `excise_synthetic_model_inputs.RData`.

The required model inputs are:

- `excise_macro_alc`
- `excise_volume_alc`
- `excise_prices_alc`
- `excise_bau_rates_alc`
- `excise_sim_rates_alc`
- `excise_vat_alc`
- `excise_tariffs_alc`
- `excise_abv_alc`
- `excise_inc_elasticities_alc`
- `excise_price_elasticities_alc`
- `excise_retail_margins_alc`
- `excise_scenarios_alc`
- `excise_base_calc_alc`
- `excise_rate_structure_alc`

The macroeconomic table contains the forecast years, GDP growth, inflation, population, population growth, and the CPI path used to convert nominal outcomes into real terms. The volume table contains product-level quantities by flow. The price table contains product-level retail prices by flow. The model uses four flows: off-trade, on-trade, domestic, and import.

The policy tables define current excise rates, simulated excise rates, VAT rates, tariffs, alcohol-by-volume values, excise base definitions, and the excise rate structure. The rate structure identifies whether the excise is ad valorem, specific by volume, or specific by alcohol content. The scenario table identifies the model horizon through scenario labels such as `t0`, `t1`, `t2`, and later years.

## Alcohol products and tax base

The current version of the module focuses on three alcohol products: Beer, Spirits, and Wine. Each product has synthetic information on volumes, prices, alcohol content, excise rates, VAT rates, tariffs, income elasticity, price elasticity, and retail margins.

The model calculates outcomes separately for off-trade, on-trade, domestic, and import flows. This distinction allows the model to capture differences between retail-market sales, domestic-market flows, and imported products. Imports also include import duties in the price decomposition.

The excise base can be defined as an ex-factory price base or as a net-of-tax base. For an ex-factory price base, the model removes VAT and retail margin from the retail price and then calculates the excise base using the excise rate. For a net-of-tax base, the model derives the base from the tax-inclusive price by removing VAT and excise.

## Excise rate structure

The Excise Module supports three excise rate structures:

- `ad_valorem`
- `specific_vol`
- `specific_abv`

For an ad valorem excise, the excise per unit is calculated as the excise base multiplied by the applicable excise rate. For a specific excise by volume, the excise per unit is the rate itself. For a specific excise by alcohol content, the excise per unit is calculated using the alcohol-by-volume value and the applicable rate.

The baseline calculation uses the current excise rate table and the baseline rate structure. The simulation calculation uses the updated simulation-rate table created by the Shiny interface. The user-selected changes are stored in `excise_sim_rates_alc_updated` and passed to the simulation calculator.

The current Shiny application allows users to enter the simulated excise rate, select the type of excise, and apply the change either only to the selected year or to all following years. In the demonstration horizon, `t0` is the base year and `t1` still uses business-as-usual rates. Simulation policy-rate changes therefore begin from `t2`.

## Baseline excise duty calculation

The baseline calculation estimates excise outcomes under the business-as-usual policy framework. It uses volumes, prices, current excise rates, VAT rates, tariffs, macroeconomic assumptions, ABV values, elasticities, retail margins, scenario information, base definitions, and the baseline excise rate structure.

The baseline script first constructs the base-year price decomposition. It calculates the excise base, excise per unit, VAT component, retail margin component, import duties for imports, and net-of-tax prices for each product and flow. It then prepares scenario-specific helper tables for later years using the macroeconomic forecast path.

For the base year, the model calculates excise per unit and total expenditure checks for each flow. For later years, the model updates excise bases, prices, volumes, and revenue outcomes using inflation, growth, elasticities, and the excise policy settings. The baseline output is used as the reference case against which simulation results are compared.

The baseline calculator produces business-as-usual outputs for volumes, prices, industry revenue, excise revenue, and real excise revenue. These outputs are later combined with the simulation outputs in the chart-preparation script.

```{r gui_excise, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/ExciseModule/GUI_Excise.png"
)
```



## Reform simulation for alcohol excise policy changes

The simulation calculation estimates the effect of user-defined alcohol excise policy changes. The Shiny app allows the user to select the product, year, excise type, and simulated excise rate. The selected changes are shown in the Selected Simulation Parameters table. When the simulation is run, the app rebuilds `excise_sim_rates_alc_updated` from the original simulation rate table and the staged user changes.

The simulation calculator uses the updated simulation-rate table while preserving the original business-as-usual rates for the baseline calculation. This separation is important because the baseline and simulation must not move together. The baseline calculator uses the original rates, while the simulation calculator uses the updated rates.

The simulation also uses behavioural assumptions. The Shiny app provides sliders for price elasticity for Beer, Spirits, and Wine, a common income elasticity, and a retail margin assumption. These values are assigned into the R session before the model scripts are sourced, so the simulation uses the current values selected by the user.

The model applies policy changes mechanically through prices, tax burdens, and elasticity-based volume responses. It does not estimate broader macroeconomic feedback, cross-product substitution beyond the modelled elasticity structure, industry strategic responses, illicit-market responses, or administrative compliance changes.

## Simulation results

The Results section of the Shiny application contains three output tables:

- Volume
- Nominal Revenues
- Real Revenues

The volume table compares business-as-usual and simulation volumes by year, product, and trade flow. It includes off-trade and on-trade volume outcomes and calculates percentage changes between the business-as-usual and simulation scenarios.

The nominal revenue table reports nominal excise revenue by product and flow. It includes off-trade, on-trade, domestic, and import excise revenue outcomes. The real revenue table reports the corresponding real excise revenue values using the CPI path from the macroeconomic input table.

The simulation result is compared with the baseline result. The fiscal effect is the difference between simulation revenue and business-as-usual revenue. Because the model also updates prices and volumes, the simulation can show both the direct effect of the rate change and the indirect effect through behavioural volume responses.

```{r excise_volume, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/ExciseModule/Volume.png"
)
```


## Dashboard and analysis outputs

The dashboard provides a visual summary of the alcohol excise simulation results. The Shiny interface includes two dashboard groups:

- Impact: Volume & Excise
- Prices: Nominal vs Real

The Impact dashboard shows excise revenue and volume outcomes over time. It compares business-as-usual and simulation results for each alcohol product. Users can switch between Beer, Spirits, and Wine using the dashboard controls.

The Prices dashboard shows nominal and real retail-price outcomes. It compares business-as-usual and simulation prices over the forecast horizon. This helps users see how the simulated excise-policy change affects retail prices and how inflation changes the interpretation of nominal price paths.

The chart-preparation script combines the business-as-usual and simulation outputs into `merged_bu_sim_alc`. This combined dataset is used by the dashboard functions and by the tables. It contains scenario labels, years, products, source indicators, volumes, prices, nominal excise revenues, and real excise revenues.

```{r excise_dashboard, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/ExciseModule/Dashboard.png"
)
```


## Current implementation scope

The current Shiny implementation is an Alcohol Excise Module. It does not currently implement tobacco or sugar-sweetened beverage calculations. It also does not currently include an excise tax expenditure table. The active outputs are volume, nominal excise revenue, real excise revenue, and dashboards for volume, excise revenue, nominal prices, and real prices.

The model is still useful for broader excise-policy demonstration because it includes the core mechanics needed for health-related excise simulation: product volumes, prices, ad valorem and specific excise structures, alcohol content, elasticity assumptions, inflation adjustments, and comparison of baseline and simulation scenarios.

## Interactive Excise Simulation Tool

A live version of the Excise Module is implemented as an interactive Shiny application. The Shiny interface allows users to select products, choose the simulation year, define excise-rate changes, select the excise structure, adjust elasticity and retail-margin assumptions, and run simulations directly through the web-based model.

After the simulation is executed, the application automatically generates volume tables, nominal excise revenue tables, real excise revenue tables, and dashboard visualizations. This live version provides a practical way to test alternative alcohol excise policy scenarios and immediately observe their fiscal, price, and volume effects.

You can access the live version of the Excise Module through the following link: https://wb-data-analytics-app.shinyapps.io/excise-module/
