```{r page_setup_vat, include=FALSE}
options(encoding = "UTF-8")
try({
  if (.Platform$OS.type == "windows") {
    Sys.setlocale("LC_CTYPE", "English_United States.utf8")
  } else {
    Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
  }
}, silent = TRUE)
```

# VAT Gap Module

The VAT Gap Module estimates potential VAT liabilities and simulates the revenue impact of changes in VAT policy parameters. The module focuses on VAT applied to goods and services by product group, industry, and use category. It uses synthetic Supply and Use Table / Input-Output Table data, VAT policy parameters by CPA product, household expenditure data, COICOP classifications, growth factors, and macro-fiscal indicators.

The module compares VAT outcomes under a baseline system and a user-defined simulation scenario. In the current Shiny application, users can select a simulation year, choose a product or industry code, adjust the exempt share, reduced-rate share, standard-rate share, preferential VAT rate, and standard VAT rate, and apply the change to the selected year or to all following forecast years. The results are presented through revenue impact tables, tax expenditure tables, VAT gap metrics, sectoral revenue charts, and household distributional outputs.

The current demonstration version of the module is based on synthetic anonymized data. The synthetic data reproduce the structure required by the model, including SUT / IOT matrices, VAT rate and taxability parameters, household expenditure data, CPA-COICOP concordance, growth factors, and macro-fiscal indicators. Because the data are synthetic, the results should be interpreted as illustrative simulation results rather than official VAT gap estimates.

## Required VAT data

The VAT Gap Module requires statistical and policy data that identify the VAT base, the applicable VAT treatment, and the potential VAT liability under current law and under reform scenarios. The core statistical data are Supply and Use Table / Input-Output Table data. In the current implementation, the main SUT / IOT inputs are supply at product-industry level, use at purchaser prices, use at basic prices, and observed VAT values.

The module also requires VAT policy parameters by CPA product or product-industry group. These policy parameters include the current exempt share, current reduced-rate share, current fully taxable share, preferential VAT rate, standard VAT rate, simulated exempt share, simulated preferential-rate share, and simulated standard-rate share. These parameters define how each product group is treated under the baseline and simulation scenarios.

Household Budget Survey and COICOP data are also used. The household data provide expenditure by household and COICOP category. The COICOP information is linked to CPA products using a CPA-COICOP concordance. This allows the module to estimate household-level VAT burdens and distributional effects by consumption decile or centile.

The module also uses forecast growth factors and macro-fiscal indicators. Growth factors are used to project VAT bases and revenues across the forecast horizon. Macro-fiscal indicators, including nominal GDP and net VAT revenue, are used to express selected outputs as a percentage of GDP or to compare modelled VAT revenue with fiscal aggregates.

For demonstration and testing purposes, the current version of the module uses synthetic data. These data reproduce the structure required by the VAT model but do not represent official national accounts, official household survey records, actual VAT returns, or a specific country.

## VAT-related goods and services

The module focuses on goods and services classified by CPA product or product-industry code. Each product group is linked to a VAT treatment under the baseline and simulation scenarios. The treatment can include exemption, reduced-rate taxation, or standard-rate taxation.

The model distinguishes between different uses of goods and services. These include intermediate consumption by businesses, household final consumption, final consumption by non-profit institutions serving households, final consumption by government, capital formation, exports, imports, trade and transport margins, and taxes less subsidies on products.

This structure is important because VAT is not calculated only from household consumption. The model also estimates VAT related to non-deductible intermediate inputs and institutional sectors where VAT may become a final cost. At the same time, exports are treated separately because they do not generate final VAT liability in the same way as domestic taxable consumption.

## Tax base: final consumption, intermediate consumption, imports, and investment

The VAT tax base is constructed from Supply and Use Table / Input-Output Table values. The model starts from product-level supply and use data, including supply, imports, trade and transport margins, taxes less subsidies on products, intermediate consumption, final household consumption, NPISH consumption, government consumption, gross fixed capital formation, changes in inventories, and exports.

The data transformation script prepares these components into product-level and industry-level tables. It calculates total supply at basic prices, total supply at purchaser prices, total intermediate consumption, final consumption, gross capital formation, total final uses, and total use at purchaser prices.

The VAT calculator then uses these transformed data together with VAT policy parameters. The benchmark VAT calculation applies a benchmark VAT rate to the relevant final demand base. The baseline calculation applies current policy treatment, including exempt shares, reduced-rate shares, and standard-rate shares. The simulation calculation applies the user-selected reform parameters.

The model separates VAT generated by businesses, households, NPISHs, and government. It also creates product-level VAT outputs, which allow results to be reported by CPA product, sector, and institutional use category.

## VAT rate structure

The simplified VAT model uses a benchmark VAT rate, a standard VAT rate, and one preferential VAT rate. The benchmark VAT rate is used to estimate potential VAT revenue under a broad benchmark VAT system. The standard VAT rate is applied to the share of each product that is fully taxable. The preferential VAT rate is applied to the share of each product taxed at the reduced rate.

For each CPA product, the policy table identifies the proportion that is exempt, the proportion taxed at the preferential rate, and the proportion taxed at the standard rate. The fully taxable share is calculated as one minus the exempt share and the reduced-rate share.

In the Shiny application, users can change the preferential VAT rate, the standard VAT rate, the exempt proportion, the preferential-rate proportion, and the standard-rate proportion for the selected product or industry. The user can also toggle simulation rates and apply the change to the selected year or to all following forecast years.

The model applies these policy parameters mechanically. It recalculates VAT liability using the updated rates and taxable proportions while keeping the underlying SUT / IOT and household expenditure data unchanged.

## Exemptions, reliefs, and special treatment

VAT exemptions and reduced rates are represented through product-level taxable proportions. For each CPA product, the model identifies the share that is exempt, the share subject to a preferential VAT rate, and the share subject to the standard VAT rate. These shares determine the effective VAT treatment applied to the relevant tax base.

The benchmark system assumes broader taxation of the VAT base at the benchmark VAT rate. The difference between benchmark VAT and estimated VAT under current policy is used to measure the policy gap and VAT-related tax expenditures. Exemptions and reduced rates therefore reduce estimated VAT revenue compared with the benchmark.

The model also includes special handling for imputed rents of owner-occupied dwellings and construction-related categories where relevant in the calculation scripts. These treatments are part of the simplified synthetic model and should be interpreted as demonstration assumptions rather than country-specific legal rules.

## Baseline VAT calculation

The baseline calculation estimates VAT revenue under the current policy framework. It uses the transformed SUT / IOT data, VAT policy parameters, observed VAT information, benchmark VAT assumptions, and calibration logic.

The model first transforms raw SUT / IOT inputs into product-level supply, use, and VAT base tables. It then calculates benchmark VAT revenue, uncalibrated VAT revenue, and calibrated VAT revenue. Benchmark VAT represents revenue under the benchmark VAT system. Uncalibrated VAT represents modelled VAT before calibration. Calibrated VAT represents the VAT result after calibration to the available VAT control.

The baseline also estimates VAT gap metrics. The total VAT gap is the difference between benchmark VAT and calibrated VAT. The policy gap reflects the revenue effect of policy provisions such as exemptions and reduced rates. The compliance gap reflects the difference between modelled potential revenue and calibrated or observed VAT revenue.

The baseline result is used as the reference case for the simulation. It provides baseline VAT revenue, VAT by product group, VAT by institutional sector, tax expenditure estimates, and VAT gap indicators.

```{r gui_vat, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/GUI_VAT.png"
)
```



## Reform simulation for VAT policy changes

The module simulates the revenue impact of changes in VAT policy parameters. Users may adjust VAT treatment for selected CPA products by changing the exempt proportion, preferential-rate proportion, standard-rate proportion, preferential VAT rate, or standard VAT rate.

The Shiny application stores selected policy changes in a simulation parameter table. When the simulation is run, the selected changes are applied to the simulation VAT policy table. The simulation then recalculates VAT revenue using the updated tax treatment. If the user selects the option to apply changes to following years, the policy change is applied from the simulation year through the remaining forecast horizon.

The simulation compares baseline VAT revenue with simulation VAT revenue. The fiscal impact is calculated as simulation VAT revenue minus baseline VAT revenue. For years before the selected simulation year, simulation values are replaced with baseline values so that the reform only affects the selected year and later years.

The current version of the module applies policy changes mechanically. It does not estimate behavioural responses such as changes in consumption, substitution between products, changes in compliance behaviour, price pass-through, or producer responses. Distributional outputs show the mechanical effect of the VAT structure on household expenditure data.

## Simulation results

The results show expected VAT revenue under baseline and simulation scenarios. The baseline result represents VAT revenue under the current policy framework. The simulation result represents VAT revenue after applying the selected reform parameters. The fiscal impact is the difference between simulation VAT revenue and baseline VAT revenue.

Results are presented in local currency units and as a percentage of GDP. They may also be disaggregated by year, CPA product, product description, institutional sector, household consumption category, and scenario.

The model reports several aggregate VAT indicators, including benchmark VAT, uncalibrated VAT, calibrated VAT, total VAT gap, policy gap, and compliance gap. These indicators help users distinguish between the benchmark tax base, policy-related revenue losses, and compliance-related gaps.

The model also separates VAT revenue generated by businesses, households, NPISHs, and government. This allows users to identify which sectors and product groups contribute most to VAT revenue and to the fiscal impact of a reform.

```{r vat_revenue, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/RevenueImpact.png"
)
```




## Dashboard and analysis outputs

The Shiny dashboard provides a visual summary of the VAT simulation results. It includes sections for policy parameter selection, VAT revenue impact, tax expenditures, and dashboard visualizations.

The policy parameter section allows users to select a simulation year, select a product-industry code and name, enter VAT rates and taxable proportions, toggle simulation rates, apply changes to future years, and add the change to the selected simulation table.

The revenue impact section reports VAT revenue projections. The tax expenditure section reports tax expenditure projections. The dashboard visualization section includes five groups of charts: Fiscal Impact, Sectoral Breakdown of VAT Revenues, Tax Expenditures, VAT-GAP Metrics, and Distributional Impacts on Households.

The Fiscal Impact dashboard shows VAT revenue forecasts, VAT revenue as a percentage of GDP, the structure of VAT by institutional sector, and a treemap by sector. The Sectoral Breakdown dashboard shows VAT revenue by CPA sector, including total VAT, VAT generated by households, VAT generated by businesses, and VAT generated by government.

The VAT-GAP Metrics dashboard shows benchmark VAT, uncalibrated VAT, calibrated VAT, policy gap, compliance gap, total VAT gap, and the structure of VAT revenues by institutional sector. The Tax Expenditures dashboard shows VAT tax expenditures over time and by CPA product. The Distributional Impacts dashboard applies effective VAT rates to household expenditure data and reports the distribution of VAT burden across household groups.

```{r vat_dashboard, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/Dashboard.png"
)
```

## Tax expenditures

The module estimates VAT-related tax expenditures by comparing benchmark VAT revenue with estimated VAT revenue under the policy system. Tax expenditures represent the revenue forgone because selected products or sectors are exempt or taxed at preferential rates instead of being taxed under the benchmark VAT system.

In the simplified VAT Module, the tax expenditure benchmark sets the VAT base under the benchmark system and compares it with the revenue estimated under current or simulated policy treatment. The difference between benchmark revenue and estimated revenue is reported as a tax expenditure. The model calculates tax expenditures at CPA product level and then aggregates them across the forecast horizon.

Tax expenditure results are presented by year and scenario. They may also be shown by CPA product or product group, in local currency units, and as a percentage of GDP. This allows policymakers to assess the fiscal cost of VAT exemptions, reduced rates, and other preferential treatments.

```{r vat_taxependitures, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/TaxExpenditures.png"
)
```

## Interactive VAT Gap Simulation Tool

The VAT Gap Module is implemented as an interactive Shiny application. The Shiny interface allows users to select VAT policy parameters, define reform scenarios, adjust VAT rates and taxable proportions, apply changes from a selected year onward, and run simulations directly through the web-based model.

After the simulation is executed, the application automatically generates VAT revenue projections, VAT gap metrics, tax expenditure estimates, sectoral revenue breakdowns, household distributional outputs, and dashboard visualizations. This live version provides a practical way to test alternative VAT policy scenarios and immediately observe their fiscal, sectoral, and distributional effects.

You can access the live version of the VAT Gap Module through the following link: https://wb-data-analytics-app.shinyapps.io/vat-module/
