# VAT Gap Module

The VAT Gap Module estimates potential VAT liabilities and simulates the revenue impact of changes in VAT policy and compliance assumptions. The module focuses on VAT applied to final consumption, intermediate consumption, investment, imports, and other taxable uses of goods and services in the economy. VAT is usually one of the most important sources of tax revenue, and VAT gap analysis can support revenue mobilization, compliance risk assessment, tax expenditure analysis, and policy reform design.

The module is usually based on statistical data rather than administrative tax data. The main data sources are Supply and Use Tables and Input-Output Tables. These tables provide a structured representation of the economy by product, industry, and use category, and allow the model to estimate the potential VAT base under the current VAT system. Because these are national accounts and statistical data sources, they do not directly reflect taxpayer declarations or actual VAT payments. Instead, they provide an economy-wide estimate of transactions that may be subject to VAT.

The calculation is also based on Household Budget Survey data. Household Budget Survey data may be available in aggregate format, for example by COICOP 4-digit category, or at a disaggregated micro level. Aggregate HBS data can be used to distribute final household consumption across VAT categories, while micro-level HBS data allows more detailed analysis by household type, income group, region, or consumption pattern. The use of HBS data is particularly important where VAT rates, exemptions, or reduced rates differ across consumption categories.

Supply and Use Tables and Input-Output Tables are not usually published every year. Depending on the country, SUTs are often published with a lag of at least three years, while IOTs may be published every five years or more frequently. This means that VAT gap modeling is often based on statistical data that is already several years old. For this reason, extrapolation is often required to update the tables to more recent years. Extrapolation may use national accounts growth rates, price indices, household consumption trends, import data, sectoral output indicators, or other macroeconomic and statistical sources.

## Required VAT data

The VAT Gap Module requires data that identifies the VAT base, the applicable VAT treatment, and the potential VAT liability under current law. The data should normally include product category, industry or activity, use category, final consumption, intermediate consumption, gross fixed capital formation, imports, exports, VAT rate, exemptions, reduced rates, zero-rating, non-deductible input VAT, and other special VAT treatments.

The core statistical data may be provided through Supply and Use Tables or Input-Output Tables. These tables should provide sufficient detail to distinguish between taxable, exempt, zero-rated, and out-of-scope transactions. Where possible, the data should identify the value of goods and services by product group and by user, including households, government, non-profit institutions, businesses, capital formation, exports, and imports.

Household Budget Survey data is also required to improve the allocation of household final consumption to VAT categories. This data may be provided in aggregate format by COICOP 4-digit category or in disaggregated micro format. When micro data is available, the module can provide additional distributional outputs, including the impact of VAT policy by household income group, expenditure group, region, or other household characteristics.

## VAT-related goods and services

The module focuses on goods and services that form part of the VAT base. These may include household consumption goods, services, imports, intermediate inputs, investment goods, and other taxable transactions. The treatment of each product or service depends on the national VAT legislation and the applicable VAT rate structure.

Goods and services may be subject to the standard VAT rate, reduced VAT rates, zero-rating, exemptions, or special treatment. Standard-rated goods and services are taxed at the main VAT rate. Reduced-rated goods and services are taxed at lower rates for policy, social, or administrative reasons. Zero-rated goods and services are taxed at a rate of zero, usually allowing input VAT recovery. Exempt goods and services are not taxed on output, but input VAT may not be recoverable, which can generate embedded VAT.

The classification of goods and services is important because each category may have a different VAT treatment, effective tax rate, compliance assumption, and policy objective. The module uses the product structure from SUTs, IOTs, and HBS data to apply the correct VAT rules and to report results by VAT category.

## Tax base: final consumption, intermediate consumption, imports, and investment

The VAT tax base may be defined in several ways depending on the structure of the VAT system and the available statistical data. In principle, VAT is intended to tax final consumption, but the actual VAT base may also include non-deductible VAT on intermediate consumption, exempt activities, government consumption, non-profit institutions, and capital formation where input VAT recovery is restricted.

For household final consumption, the tax base is usually estimated using SUT or IOT product-level consumption values, complemented by HBS data. HBS data helps allocate consumption to detailed categories and apply the correct VAT treatment. If HBS data is available only in aggregate format, the model may use COICOP 4-digit categories. If micro-level HBS data is available, the model may estimate VAT liabilities at household level before aggregating the results.

For intermediate consumption and investment, the model identifies sectors or activities where input VAT is not fully deductible. This is especially important for exempt sectors such as financial services, health, education, real estate, or public services, depending on national VAT law. In these cases, VAT on inputs may become a final cost and therefore part of the potential VAT base.

Imports may also be included in the VAT base, especially where imported goods and services are consumed domestically or used by exempt sectors. Exports are generally zero-rated and are therefore excluded from final VAT liability, although they may remain relevant for input VAT recovery and refund analysis.

## VAT rate structure

VAT systems can be structured with a standard rate, one or more reduced rates, zero-rating, exemptions, and special regimes. The standard VAT rate is the main rate applied to most taxable goods and services. Reduced rates are lower rates applied to selected goods or services, often for social, distributional, or policy reasons. Zero-rating applies a VAT rate of zero while normally allowing input VAT deduction. Exemptions remove output VAT but may restrict input VAT recovery.

Some VAT systems may also include special schemes for small taxpayers, agriculture, public bodies, financial services, real estate, margin schemes, or simplified regimes. These provisions can affect the effective VAT base and the amount of VAT that would be collected under current law.

The module applies the relevant VAT rate or treatment to each product, use category, or household consumption category. Where the statistical classification does not perfectly match the legal VAT classification, mapping assumptions are required. These assumptions link SUT, IOT, or COICOP categories to the applicable VAT treatment.

## Exemptions, reliefs, and special treatment

VAT systems often include exemptions, reduced rates, zero-rating, thresholds, special schemes, and other preferential treatments. These provisions may apply to specific goods, services, sectors, institutions, taxpayers, or transactions. For example, special treatment may apply to basic food, medicines, education, health services, financial services, residential rent, agriculture, small businesses, exports, or public sector activities.

The module uses this information to distinguish between the benchmark VAT liability and the actual VAT liability under current law. This allows users to estimate the fiscal cost of exemptions, reduced rates, zero-rating, and special VAT treatments. It also allows users to simulate reforms that remove, reduce, expand, or modify these provisions.

## Baseline VAT calculation

The baseline calculation estimates potential VAT revenue under the current policy framework. It uses the observed or extrapolated VAT base from SUTs, IOTs, HBS data, and other statistical sources, together with the current VAT rates and treatment rules. The baseline represents the potential VAT liability under current law before estimating the VAT gap.

For standard-rated and reduced-rated goods and services, VAT is calculated by multiplying the taxable base by the applicable VAT rate. For zero-rated goods and services, the output VAT liability is zero, but input VAT recovery may still be considered depending on the structure of the model. For exempt goods and services, output VAT is not charged, but non-deductible input VAT may be estimated where relevant.

If SUTs or IOTs are not available for the target year, the model extrapolates the base to more recent years. This may involve updating product-level values using growth rates, price indices, household consumption data, sectoral output data, import data, or other macroeconomic indicators. Extrapolation is necessary because Supply and Use Tables and Input-Output Tables are often published with significant delays and may not be available annually.


```{r gui_vat, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/GUI_VAT.png"
)
```



## Reform simulation for VAT policy changes

The module can simulate the revenue impact of changes in VAT policy. Users may test increases or reductions in the standard VAT rate, changes in reduced rates, removal of exemptions, expansion of zero-rating, introduction of new reduced rates, changes in the VAT treatment of specific products, or reforms to special VAT regimes.

The simulation compares the current VAT burden with the proposed VAT burden. Where appropriate, the module may estimate the effect of the policy change on the VAT base using behavioural assumptions. These assumptions may reflect changes in consumption, substitution between products, price effects, or household responses to VAT rate changes.

If micro-level Household Budget Survey data is available, the module may also simulate the distributional effects of VAT reforms across households. This allows users to assess how VAT policy changes affect different income groups, expenditure groups, regions, or household types. If only aggregate HBS data is available, distributional analysis may be more limited, but the data can still support improved allocation of consumption across VAT categories.

After the simulated VAT base is estimated, the module calculates the simulated VAT liability using the proposed policy parameters. The revenue impact is calculated as the difference between simulated VAT revenue and baseline VAT revenue.

## Simulation results

The results show the expected VAT revenue under current law and under the simulation scenario. The current law result represents the baseline potential VAT revenue that would be collected if the current VAT system were fully applied to the estimated VAT base. The simulation result represents the potential VAT revenue expected after applying the proposed policy changes. The fiscal impact is the difference between the two.

Results may be presented in nominal terms, such as local currency units, and as a percentage of GDP. They may also be disaggregated by year, product category, COICOP category, industry, use category, VAT rate, VAT treatment, household group, or sector. This allows users to identify which parts of the VAT base generate the largest revenue effects.

The module may also compare potential VAT revenue with actual VAT collections. The difference between potential VAT revenue and actual VAT revenue represents the estimated VAT gap. This gap may reflect non-compliance, fraud, administrative weaknesses, timing differences, policy gaps, measurement issues, or limitations in the statistical data.

```{r vat_revenue, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/RevenueImpact.png"
)
```




## Dashboard and analysis outputs

The dashboard provides a visual summary of the VAT simulation results. It may include summary boxes for potential VAT revenue, actual VAT revenue, VAT gap, simulation revenue, and fiscal impact. It may also include charts showing the evolution of VAT revenue over time, the distribution of potential VAT revenue by product group, and the comparison between current law and simulation results.

Additional outputs may show the VAT base by product, COICOP category, industry, use category, VAT rate, or VAT treatment. The dashboard may also include estimates of non-deductible input VAT, VAT by household group, revenue by rate category, and the fiscal cost of exemptions, reduced rates, and zero-rating. These outputs help users understand not only the total revenue effect, but also the distribution of the impact across goods, services, sectors, and households.


```{r vat_dashboard, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/Dashboard.png"
)
```

## Tax expenditures

The module can also estimate VAT-related tax expenditures. These represent the revenue forgone because of exemptions, reduced rates, zero-rating, thresholds, or special treatments compared with a benchmark VAT system.

Tax expenditure results may be presented by year, product group, COICOP category, industry, VAT treatment, household group, or sector. They may be shown in local currency units and as a percentage of GDP. This allows policymakers to assess the fiscal cost of preferential VAT provisions and evaluate whether these provisions remain justified from a revenue, social, distributional, or economic perspective.


```{r vat_taxependitures, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  "https://raw.githubusercontent.com/jordans78/tax_model_website/main/img/VATModule/TaxExpenditures.png"
)
```

## Interactive VAT Gap Simulation Tool

A live version of the VAT Gap Module can be implemented as an interactive Shiny application. The Shiny interface allows users to select VAT rates, define reform scenarios, adjust VAT treatment rules, apply extrapolation assumptions, and run simulations directly through the web-based model. After the simulation is executed, the application automatically generates potential VAT revenue estimates, VAT gap estimates, tax expenditure estimates, distributional outputs where HBS micro data is available, and dashboard visualizations.

This live version provides a practical way to test alternative VAT policy scenarios and immediately observe their fiscal, sectoral, and distributional effects.You can access the live version of the Excise Module through the following link:https://wb-data-analytics-app.shinyapps.io/vat-module/
