---
title: "Tax Modeling Toolkit"
author: "World Bank Tax Team"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook:
    css: www/style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

source("R/01_load_packages.R")
source("R/02_prepare_data.R")
source("R/03_create_tables.R")
source("R/04_create_charts.R")
```

# Introduction

This website presents an example of an HTML documentation page created with R.

The website can be used to document a tax modeling toolkit covering:

- VAT
- PIT
- CIT
- Customs duties
- Excise duties

The website is static HTML, but it can include R tables, R charts, formulas, and links to Shiny applications.

## Main purpose

The purpose of the toolkit is to provide a structured framework for:

- preparing tax data;
- defining baseline and reform scenarios;
- estimating revenue impacts;
- presenting simulation results;
- documenting assumptions and methodology.

## Example data

```{r}
head(tax_results)
```
